# Aside: Understanding Methods and Generics in R {- .aside-chapter}

## You've Picked Up a Lot of R

You can wrangle data, fit models, make plots, write functions. But you might also be starting to wonder what's actually going on under the hood.

For example, how does `summary()` know what to do when you give it the output from `lm`? Or why does `plot()` sometimes give you a histogram and other times a scatterplot?

A good way to start answering those questions is to learn about methods and generic functions.

## Why Use Generics?

It's nice when a function does the right thing depending on the kind of object you give it. You don't want to memorize a different function name for every kind of model, you just want to call `summary()`, `plot()`, or `print()` and get something useful. That's what generic functions are for.

In R, many common functions like `summary()`, `plot()`, and `print()` are generic. They automatically call the version that matches the class of your object.

Let's run a linear model and use that as an example. We will look at miles per gallon and vehicle weight from the `mtcars` dataset.

```{r}
#| label: generics-fit-lm
fit <- lm(mpg ~ wt, data = mtcars)
```

Ok. We have a new object called `fit` that is a `list` with a bunch of stuff in it that the authors of `lm` think will be useful to you. Look at its structure to see what's in there.

```{r}
#| label: generics-str
str(fit)
```

That list has 12 things in it including the coefficients, residuals, fitted values, and a lot more. But getting the information you want in terms of interpreting the linear model isn't obvious. The object `fit` doesn't directly contain the $R^2$ or p-values. Those are computed when you call `summary(fit)`.

```{r}
#| label: generics-summary-lm
summary(fit)
```

Check that out. We got a nicely formatted output that helps us interpret the linear model. But note that the output of `summary(fit)` is very different from getting a summary of a numeric vector like the `mpg` column that went into the model.

```{r}
#| label: generics-summary-vector
summary(mtcars$mpg)
```

### Same Function, Different Behavior. What Gives?

We used the same function, `summary`, and got two very different results. The answer is generics and methods.

When we called `summary(fit)`, we weren't calling some one-size-fits-all summary function. We were calling a specific method written for objects of class `"lm"`.

## How This Works

Here's what R does when you call `summary(fit)`:

1. It checks the class of the object (`class(fit)`).
2. It looks for a function named `summary.classname`, in this case `summary.lm`.
3. If it finds that method, it runs it.
4. If not, it falls back to `summary.default()`.

You can look under the hood:

```{r}
#| label: generics-dispatch-lookup
class(fit)
methods(summary)
summary.lm
getS3method("summary", "lm")
```

## What's the Difference Between a Generic and a Method?

A generic function is the function you call, like `summary()` or `plot()`. It doesn't actually do the work itself. Instead, it decides which method to run based on the class of the object.

A method is the specific function written for a particular class. `summary.lm()` handles linear models, `summary.data.frame()` handles data frames, and `summary.default()` is the fallback if no class-specific method is found.

You can look at the generic directly:

```{r}
#| label: generics-generic-body
summary
```

That call to `UseMethod("summary")` tells R to dispatch to the right method depending on the class of the object.

## Common Generic Functions

Here are some other generic functions you're likely to run into:

- `print()`: display a human-readable version
- `summary()`: provide a richer summary
- `plot()`: show plots
- `predict()`: make predictions
- `residuals()`: return residuals
- `fitted()`: return fitted values
- `coef()`: extract coefficients
- `update()`: refit a model with changes
- `AIC()` / `BIC()`: compare model fit
- `anova()`: compare nested models
- `terms()` / `formula()`: extract parts of model structure

## How to Explore and Understand Methods

Want to know what's going on behind a function call? Here are a few tools:

```{r}
#| label: generics-explore-methods
methods(summary)
methods(plot)
methods(class = "lm")
getS3method("summary", "lm")
class(fit)
```

## Where This Shows Up

You already ran into this once, quietly, back in the time chapter: `plot(LakeHuron)` and `co2`'s calendar-grid printout both work because `plot` and `print` dispatch on the object's class. You'll run into it again in the regression chapter, where `summary()` on an `lm` object and `summary()` on a `gls` object hand you different tables built from different assumptions, and knowing that you're calling two different methods, not one function with two moods, is the whole point of reading either one carefully. Whenever this book says a function "knows" what to do with an object, this is the mechanism doing the knowing.
