# Regression

```{r}
#| label: reg-setup
#| echo: false
set.seed(872)
```

## Big Idea

The last chapter ended by pointing here. Once a cross-correlation convinces you that one series tracks another, the natural next move is to fit a model and read the slope: how much does $y$ change when $x$ does? That is regression, and you have done it since intro stats. The wrinkle is that our data are ordered in time, and the errors of a time-series regression are almost never the tidy independent draws that ordinary least squares assumes. They carry memory. Yesterday's error tells you something about today's.

That memory is the saboteur from the start of the book wearing a regression costume. When the residuals of a regression are autocorrelated, you have broken the independence assumption that the standard errors, the $t$-statistics, and the $p$-values all rest on. The estimate of the slope can still be fine. What goes wrong is your sense of how sure you are. Worse, the trouble is sharpest in the single most common thing an environmental scientist asks of a time series: is there a trend? Fit a line against time with autocorrelated errors and the software will hand you a slope and a tiny $p$-value whether or not anything is actually changing. This chapter is about spotting that, understanding why it happens, and fixing it with generalized least squares so that the number you report is one you can stand behind.

## Packages

We stay in the tsibble idiom and read example data from CSV. The new tool is `gls` from `nlme` [@R-nlme], which fits a linear model while letting the errors carry a correlation structure. We use `forecast` [@R-forecast] for `ggAcf`/`ggPacf`, `ggridges` for the simulation figure, and `PNWColors` for the book's palette.

```{r}
#| label: reg-packages
#| message: false
library(tidyverse)
library(nlme)
library(forecast)
library(ggridges)
library(PNWColors)

sfPal <- pnw_palette("Starfish", n = 5, type = "discrete")
```

## A Trend That Isn't There

Let's start with the most dangerous case and see the saboteur in the open. We will build a series with no trend at all: pure AR(1) noise, wandering around a flat mean of zero. Then we will fit a straight line against time and ask whether the slope is significant. We know the right answer is no. There is no trend. We built it that way.

```{r}
#| label: reg-trend-plant
set.seed(22)
n <- 120
phi <- 0.85
y <- as.numeric(arima.sim(model = list(ar = phi), n = n))
time <- 1:n
```

Always look at what you built. An AR(1) series with a coefficient near one wanders. It sets off in some direction and stays there for a while before turning, because each value is mostly a copy of the one before it. To the eye it can look exactly like a series that is trending.

```{r}
#| label: reg-trend-plot
tibble(time, y) |>
  ggplot(aes(time, y)) +
  geom_line(color = sfPal[1]) +
  geom_smooth(method = "lm", se = FALSE, color = sfPal[5], linewidth = 0.7) +
  labs(x = "Time", y = "y",
       title = "A series with no trend, and the line OLS fits to it") +
  theme_minimal()
```

The red line is what ordinary least squares draws through it. Now fit that regression and look at what the software says.

```{r}
#| label: reg-trend-ols
ols_trend <- lm(y ~ time)
summary(ols_trend)
```

The $p$-value on the slope is around three in a billion. By any conventional standard that is a screaming, unambiguous trend. And it is completely false. There is no trend in this series. The only structure is the AR(1) memory, and that memory has fooled the regression into mistaking a slow random wander for a directed climb.

This is not a fluke of one unlucky seed. Let's run it many times. For a range of AR(1) strengths, we will generate a trend-free series, fit the line, and record how often the slope comes back significant at the usual 5% level. If the test were behaving, that rate would sit near 5% no matter how much memory the errors carry.

```{r}
#| label: reg-trend-mc
false_positive_rate <- function(phi, n = 100, sims = 1000) {
  hits <- replicate(sims, {
    y <- as.numeric(arima.sim(model = list(ar = phi), n = n))
    p <- summary(lm(y ~ seq_len(n)))$coefficients[2, 4]
    p < 0.05
  })
  mean(hits)
}

phis <- c(0, 0.3, 0.5, 0.7, 0.9)
fp <- tibble(phi = phis,
             rate = map_dbl(phis, false_positive_rate))
fp
```

```{r}
#| label: reg-trend-fp-plot
ggplot(fp, aes(factor(phi), rate)) +
  geom_col(fill = sfPal[5], width = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = expression("AR(1) coefficient of the errors  " * (phi)),
       y = "Trends declared significant",
       title = "OLS finds a trend in trend-free data",
       subtitle = "Each bar: share of 1000 flat AR(1) series where OLS calls the slope significant") +
  theme_minimal()
```

The dashed line is where a well-behaved test should sit. At $\phi = 0$, when the errors are independent, OLS behaves: it cries trend about 5% of the time, exactly the false-positive rate you signed up for. But as the memory in the errors climbs, so does the deception. By $\phi = 0.9$, ordinary least squares declares a significant trend in roughly two of every three series that have no trend at all. The software prints a trend either way. The only thing that changes is how often you get fooled.

## Why It Happens

The mechanism is worth getting straight, because it tells you exactly when to worry. The OLS slope estimate is still unbiased here. Across all those simulations it averages to zero, the truth. What breaks is the standard error. OLS computes the uncertainty on the slope under the assumption that every observation is an independent piece of evidence. When the errors are autocorrelated, consecutive points are partly copies of each other, so you have less independent information than the sample size suggests. OLS does not know this. It counts all $n$ points as full evidence, the standard error comes out too small, the $t$-statistic too big, and the $p$-value too optimistic.

There is a clean way to see when this bites. The bias in the OLS standard error is, to a good approximation, proportional to the sum over lags of the error autocorrelation times the predictor autocorrelation:

$$\text{SE bias} \;\propto\; \sum_k \rho_\epsilon(k)\,\rho_x(k)$$

If the predictor has no time structure of its own, every $\rho_x(k)$ is zero, that sum collapses, and the OLS standard error is about right even when the errors are autocorrelated. There is nothing for the memory in the errors to align with. But the moment the predictor carries its own autocorrelation, both terms are positive, the sum grows, and the standard error shrinks below where it should be. And the predictor with the most autocorrelation imaginable is time itself: $1, 2, 3, \dots$ climbs as smoothly as a variable can. That is why fitting a trend is the worst case, and why it is the one to be most suspicious of.

So the diagnosis is always the same. Fit your model, then look at the residuals the way we have looked at every series in this book. The ACF and PACF of the OLS residuals from our trend-free series tell the story immediately.

```{r}
#| label: reg-trend-resid
#| fig-height: 3
ggAcf(residuals(ols_trend)) + labs(title = "ACF of OLS residuals") +
  theme_minimal()
```

The slow decay is the AR(1) signature. The residuals are not independent, the independence assumption is broken, and any $p$-value from that fit is not to be trusted. The plot is the warning. Now we fix it.

## Generalized Least Squares

When the OLS assumptions are met, independent errors, equal variance, normality, OLS is hard to beat. The estimators are BLUE: Best Linear Unbiased Estimators. And OLS has a nice teaching property: you can derive it with algebra and a little calculus, without any matrix notation. That is probably how you learned it.

The problem, as we have just seen, is that time-series residuals are often autocorrelated. That violates the independence assumption, which means the standard errors on your coefficients can be badly wrong, typically too small, which makes your $p$-values too optimistic. You think you have more evidence than you actually do.

Generalized least squares is the natural fix. The difference between OLS and GLS is in what they assume about the errors. OLS assumes the errors are independent and identically distributed. GLS allows them a richer structure: a covariance matrix $\boldsymbol{\Sigma}$ that can encode both unequal variances and correlations between observations. OLS minimizes the sum of squared distances between observed and predicted values, treating every observation as equal evidence. GLS minimizes those same distances weighted by the covariance structure of the residuals, so observations whose errors are strongly correlated with their neighbors get less weight. They are carrying redundant information, and GLS knows it.

The two estimators sit side by side like this:

$$\hat{\boldsymbol{\beta}}_{OLS} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y}$$
$$\hat{\boldsymbol{\beta}}_{GLS} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{y}$$

GLS is OLS with $\boldsymbol{\Sigma}^{-1}$ inserted as a weighting matrix. When $\boldsymbol{\Sigma}$ is just $\sigma^2 \mathbf{I}$, equal variances and no correlation, the two formulas are identical and OLS is a special case of GLS. If the matrix notation is unfamiliar, the [OLS via Algebra and Matrices aside](08xAsideOLS-Algebra-Matrix.html) walks through where both formulas come from and why they are built the way they are.

In practice you do not know $\boldsymbol{\Sigma}$ exactly. With `gls()` from `nlme` you specify the *structure* of the errors, for example that they follow an AR(1) process, and the software estimates the parameters of that structure from the data by maximum likelihood or restricted maximum likelihood (REML). This is also a reason not to reach for GLS reflexively. If you meet the OLS assumptions, use OLS. If you do not, GLS is a good option, but it asks you to specify the correlation structure correctly. Get that wrong and you can do more harm than good.

Let's fix the trend regression. We fit the same model, $y$ against time, but tell `gls` that the errors follow an AR(1) process with `corAR1()`.

```{r}
#| label: reg-trend-gls
gls_trend <- gls(y ~ time, correlation = corAR1())
summary(gls_trend)
```

The slope estimate shrinks toward zero, and the standard error more than quadruples. The $p$-value climbs from three-in-a-billion to about 0.6. GLS has accounted for the memory in the errors, recognized that 120 autocorrelated points are worth far fewer than 120 independent ones, and reported the uncertainty correctly. The trend that OLS was certain of has evaporated, which is exactly what should happen, because we built a series with no trend.

## Recovering a Slope That Is There

The trend case is the saboteur at its most destructive, because the predictor and the errors share so much structure. But GLS is not only a tool for tearing down false trends. It also recovers true relationships more reliably when the errors are autocorrelated. Let's build a case where there is an effect to find.

We will make `x` ordinary white noise and build `y` as a function of `x` plus AR(1) noise. We know the true slope, so we can check who recovers it.

```{r}
#| label: reg-plant
set.seed(47)
n <- 100
phi <- 0.8
x <- rnorm(n)
epsilon <- as.numeric(arima.sim(model = list(ar = phi), n = n))
epsilon <- epsilon - mean(epsilon)
B0 <- 0
B1 <- 0.5
y <- B0 + B1 * x + epsilon
```

Notice that here the predictor `x` is white noise, so by the rule from a moment ago the OLS standard errors will be roughly right. What suffers instead is efficiency: the autocorrelated errors make the OLS slope estimate noisy, and on this particular draw it lands well off the truth.

```{r}
#| label: reg-plant-ols
ols_plant <- lm(y ~ x)
coef(ols_plant)[2]
```

The true slope is 0.5 and OLS returned something nearer 0.2. The residuals, predictably, are autocorrelated.

```{r}
#| label: reg-plant-resid
#| fig-height: 3
ggAcf(residuals(ols_plant)) + labs(title = "ACF of OLS residuals") +
  theme_minimal()
```

Before we hand this to `gls`, it is worth seeing what GLS does under the hood, because you already met the idea last chapter. Prewhitening. If we knew the AR(1) coefficient of the errors, we could transform both `y` and `x` to strip the memory out, then run an ordinary regression on the cleaned series. Estimate the coefficient from the OLS residuals, then subtract its lagged contribution from each variable.

```{r}
#| label: reg-plant-byhand
ar1 <- ar(residuals(ols_plant))$ar[1]
y_lag1 <- dplyr::lag(y)
x_lag1 <- dplyr::lag(x)
y_white <- y - ar1 * y_lag1
x_white <- x - ar1 * x_lag1

ols_white <- lm(y_white ~ x_white)
coef(ols_white)[2]
```

That slope is much closer to the true 0.5, and the residuals of `ols_white` are clean. This hand transform is the old Cochrane-Orcutt idea, and it is essentially what GLS automates, except GLS estimates the correlation structure and the regression coefficients together rather than in two passes. Here is the `gls` version.

```{r}
#| label: reg-plant-gls
gls_plant <- gls(y ~ x, correlation = corARMA(p = 1))
coef(gls_plant)[2]
summary(gls_plant)$tTable
```

We ask for `corARMA(p = 1)`, which is an AR(1) error structure. Had the residuals looked like an ARMA(1,1) we could try `corARMA(p = 1, q = 1)`, and `gls` would estimate those parameters by numerical optimization, which can fail when the model is too complex. The GLS slope lands near 0.5, the value we built in, and its standard error reflects the true uncertainty. When you inspect GLS residuals for autocorrelation, ask for the `normalized` ones, which are the residuals transformed by the estimated correlation structure and so are what should look like white noise if the structure was right.

```{r}
#| label: reg-plant-gls-resid
#| fig-height: 3
ggAcf(residuals(gls_plant, type = "normalized")) +
  labs(title = "ACF of normalized GLS residuals") +
  theme_minimal()
```

Clean.

## One Run Is an Anecdote

A single simulation proves nothing. OLS got unlucky on that seed, and you could fairly object that another draw might flip it. So let's run the whole thing thousands of times across a grid of sample sizes and error strengths, fit both OLS and GLS to each, and look at the distribution of the slope estimates. The truth is always 0.5.

```{r}
#| label: reg-mc-sim
sim_one <- function(n, phi) {
  x <- rnorm(n)
  epsilon <- as.numeric(arima.sim(model = list(ar = phi), n = n))
  epsilon <- epsilon - mean(epsilon)
  y <- 0.5 * x + epsilon
  ols <- lm(y ~ x)
  gls_fit <- tryCatch(gls(y ~ x, correlation = corARMA(p = 1)),
                      error = function(e) NULL)
  if (is.null(gls_fit)) return(NULL)
  tibble(OLS = coef(ols)[2], GLS = coef(gls_fit)[2])
}

set.seed(872)
grid <- expand_grid(n = c(50, 100, 150, 200),
                    phi = c(0.3, 0.5, 0.7, 0.9),
                    rep = 1:225)

sim <- grid |>
  mutate(out = map2(n, phi, sim_one)) |>
  unnest(out)
```

```{r}
#| label: reg-mc-ridges
#| fig-height: 5
sim |>
  pivot_longer(c(OLS, GLS), names_to = "model", values_to = "B1hat") |>
  filter(between(B1hat, 0, 1)) |>
  mutate(model = factor(model, levels = c("OLS", "GLS")),
         phi_lab = factor(phi, labels = paste0("phi == ", c(0.3, 0.5, 0.7, 0.9))),
         n_lab = factor(n)) |>
  ggplot(aes(x = B1hat, y = n_lab, fill = model)) +
  geom_density_ridges(alpha = 0.7, scale = 1.1) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  facet_grid(rows = vars(model), cols = vars(phi_lab),
             labeller = labeller(phi_lab = label_parsed)) +
  scale_fill_manual(values = c(OLS = sfPal[5], GLS = sfPal[1]), guide = "none") +
  labs(x = expression(hat(beta)[1]), y = "Sample size",
       title = "GLS recovers the true slope more tightly than OLS",
       subtitle = expression(y[t] == 0.5 * x[t] + epsilon[t] *
                             "  with " * epsilon * " an AR(1) process; dashed line is the truth"),
       caption = "3600 simulations") +
  theme_minimal() +
  theme(legend.position = "none")
```

Read down the columns as the error memory strengthens from left to right. Both methods are centered on the truth, so neither is biased. But the GLS distributions are tighter, and the gap widens as $\phi$ grows and as the sample size shrinks. At $\phi = 0.9$ with only 50 points, the OLS estimate is all over the place while GLS stays comparatively concentrated on 0.5. That is the efficiency gain. Even in this friendly case, where the white-noise predictor keeps the OLS standard errors roughly right, you still get a better estimate by modeling the structure in the errors. And in the unfriendly case, the trend, modeling the structure was the difference between a true result and a false one. Either way the lesson is the same one the whole book keeps arriving at: the dependence is there whether you account for it or not, so account for it.

## An Observed Series: Colorado River Flow

Simulations are where you check that a method does what it claims. Observed data is where you find out whether the problem shows up in the wild. A short and excellent [paper](http://onlinelibrary.wiley.com/doi/10.1002/2015GL067613/full) by Connie Woodhouse and colleagues [@Woodhouse2016] looks at how annual flow on the Colorado River responds to precipitation, soil moisture, and temperature. The river is the water supply for forty million people, so whether its flow is predictable from climate, and whether warming is bending that relationship, is not an academic question. I have put their data in `data/woodhouse.csv`.

```{r}
#| label: reg-flow-data
#| message: false
flow <- read_csv("data/woodhouse.csv") |>
  mutate(MAF = LeesWYflow / 1e6)

ggplot(flow, aes(Year, MAF)) +
  geom_line(color = sfPal[1]) +
  labs(y = "Flow (millions of acre feet)", x = "Water year",
       title = "Colorado River at Lees Ferry") +
  theme_minimal()
```

The flow record is the annual total at Lees Ferry, where the river is measured before the big downstream diversions. We will model it as a function of October-to-April precipitation, the cool-season moisture that feeds the spring runoff. Start with OLS, then look at the residuals.

```{r}
#| label: reg-flow-ols
ols_flow <- lm(LeesWYflow ~ OctAprP, data = flow)
summary(ols_flow)
```

Precipitation explains about two thirds of the variance in annual flow, which is a strong relationship and makes physical sense: more cool-season precipitation, more water in the river. But this is a time series, so the slope and its standard error are only trustworthy if the residuals are independent.

```{r}
#| label: reg-flow-resid
#| fig-height: 3
ggAcf(residuals(ols_flow)) + labs(title = "ACF of OLS residuals, flow on precipitation") +
  theme_minimal()
```

They are not. The lag-1 autocorrelation is around 0.4 and clears the significance band, so consecutive years' residuals are related. A wet year tends to follow a wet year, soil moisture and snowpack carry over, and whatever the precipitation predictor misses persists from one year to the next. The independence assumption is broken, and the OLS standard errors are suspect. Refit with an AR(1) error structure.

```{r}
#| label: reg-flow-gls
gls_flow <- gls(LeesWYflow ~ OctAprP, data = flow,
                correlation = corAR1())
summary(gls_flow)
```

The slope stays large and overwhelmingly significant, so the headline finding survives: precipitation is a strong driver of flow. Now think back to the rule from "Why It Happens." The residuals here are strongly autocorrelated, but the predictor is not. Precipitation from one year to the next is close to independent, with a lag-1 autocorrelation near zero. That puts us in the efficiency regime, not the standard-error-deflation one, so the OLS standard error was not badly inflated to begin with, and GLS actually tightens it a little rather than widening it. Either way the effect is far too strong for the autocorrelation to change the verdict. The value of running the GLS fit is that you confirmed that, rather than assuming it. You looked at the residuals, saw the memory, and checked that it did not undermine the result.

## When the Residuals Are Clean

It would be easy to come away thinking every time-series regression needs GLS. It does not. The rule is to diagnose, not to assume. Here is a case that looks like it should be trouble and turns out fine. The `data/ny_air_quality.csv` file holds daily measurements of ground-level ozone in New York over a summer, along with the weather. Ozone is a photochemical pollutant: it cooks up on hot, sunny, stagnant days. So we expect temperature to predict it, and we expect the data to be autocorrelated, because weather comes in multi-day spells. Hot days cluster.

```{r}
#| label: reg-ozone
#| message: false
air <- read_csv("data/ny_air_quality.csv")
ols_ozone <- lm(ozone ~ temperature, data = air)
summary(ols_ozone)$coefficients
```

Temperature is a strong predictor, as expected. The day-to-day persistence of the weather means temperature itself is autocorrelated. By the instinct we have built, that should set off alarms. So check the residuals.

```{r}
#| label: reg-ozone-resid
#| fig-height: 3
ggAcf(residuals(ols_ozone)) + labs(title = "ACF of OLS residuals, ozone on temperature") +
  theme_minimal()
```

Nothing. The residual autocorrelation is negligible and stays inside the band. Here is the resolution of the apparent paradox: it is fine for the predictor `x` or the response `y` to be autocorrelated. What has to be independent is the residuals. Temperature soaked up the persistence. Because the hot spells that drive ozone are themselves carried by temperature, once temperature is in the model there is no leftover day-to-day memory for the errors to inherit. The independence assumption holds, and OLS stands. You did not need GLS, and the only way you could know that was to look. The plot is what earns you the right to keep the simpler model.

## Wrapping Up

Time-series regression is the saboteur from the opening chapters in its most practical form. The slope estimate usually survives autocorrelated errors, but the uncertainty around it does not, and the damage is worst exactly where it matters most: testing for a trend against time, where the predictor's own structure aligns with the memory in the errors and OLS will conjure significance out of a flat random walk. The discipline is the same one the book has asked for throughout. Fit the model, then look at the residuals. If they are clean, as with the ozone, OLS is doing its job and you leave it alone. If they carry memory, as with the flow and the false trend, GLS lets you specify that structure and get a trustworthy answer back.

There is a non-parametric version of this same story still to tell. When people test environmental series for a trend they often reach for the Mann-Kendall test and the Theil-Sen slope precisely because those methods avoid the straight-line assumption. But Mann-Kendall assumes independence too, and on autocorrelated data it over-rejects in exactly the way OLS just did, printing trends that are not there. The fix, prewhitening the series before the test, is the same move we made by hand above. That section is coming.

## Exercises

### Push the simulation

Go back to the slope-recovery simulation and change `B1`, `phi`, and `n` by hand to get a feel for when OLS and GLS diverge. What happens to the gap between them when you make the true slope smaller relative to the noise? When does OLS get away with ignoring the autocorrelation, and when does it cost you? Then change the predictor from white noise to something with its own memory, for example `x <- as.numeric(arima.sim(list(ar = 0.8), n = n))`, refit both, and watch the OLS standard error start to lie the way it did in the trend example.

### Colorado River, with temperature

The Woodhouse paper's actual contribution is about temperature. Their point is that warm years amplify drought: "Different combinations of temperature, precipitation, and soil moisture can result in flow deficits of similar magnitude, but recent droughts have been amplified by warmer temperatures." Fit a multiple regression of flow on both October-to-April precipitation (`OctAprP`) and March-to-July temperature (`MarJulT`). Does adding temperature improve the model? Look at the residuals: are they still autocorrelated once temperature is in? Refit with GLS and compare the temperature coefficient and its standard error to the OLS version. Given your simulation work, do you think Woodhouse and colleagues are on safe footing reading a temperature effect from this record?

## Further Reading

- @Cowpertwait2009, chapter 5, covers regression with autocorrelated errors and the GLS approach, including the generalized least squares estimator and its use on environmental series.
- @Pinheiro2000 is the definitive reference for the `nlme` package, written by its authors. It is the place to go for what the correlation structures are and how the fitting actually works.
- @Zuur2009 is a gentler, ecology-facing treatment of mixed models and GLS, with worked environmental examples and a lot of practical advice about diagnosing and choosing correlation structures.
