# Aside: Cross-Validation with Time Series {- .aside-chapter}

```{r}
#| label: cv-setup
#| echo: false
#| message: false
library(tidyverse)
set.seed(872)
```

## Why Are We Doing This?

Both validation schemes used so far in this book amount to the same simple move. The forecasting chapter withheld the last five years of a river's flow, fit on what was left, and forecast across the gap. The reconstruction chapter split the instrumental record in two, calibrated on one half, and verified on the other. Both are sound, as far as they go. Both are also exactly one split: a single train, a single test, and the whole judgment about whether a model is any good rests on that one division of the data.

In most of statistics, when one split is not enough, you reach for cross-validation: divide the data into several folds, hold each one out in turn, average the errors. That machinery is standard and well understood. It is also, applied carelessly, one of the easiest ways to fool yourself with a time series. This aside is about why the textbook version breaks, what to do instead, and a surprising case where the textbook version turns out to be fine after all.

## Why You Can't Just Shuffle and Split

Ordinary k-fold cross-validation works by shuffling the rows of your data, splitting them into k roughly equal groups, training on k-1 of them, and testing on the one left out, rotating through all k. The shuffle is the part that breaks. It assumes every row is exchangeable with every other row, that swapping two observations changes nothing about the problem. For a time series that assumption is false by construction. Yesterday's flow is not exchangeable with flow from a random year a century away; it is the single best piece of information you have about today's flow, which is the entire reason autocorrelation exists in the first place.

Here is what goes wrong in practice. Build a series with plenty of memory and no usable structure beyond that memory, an AR(1) with a high coefficient, and try to predict it from nothing but its own position in time, using a method flexible enough to lean on nearby points.

```{r}
#| label: cv-leakage-sim
n_sim <- 240
y <- as.numeric(arima.sim(list(ar = 0.95), n = n_sim))
x <- 1:n_sim

knn_predict <- function(train_x, train_y, test_x, k = 15) {
  sapply(test_x, function(tx) {
    idx <- order(abs(train_x - tx))[1:min(k, length(train_x))]
    mean(train_y[idx])
  })
}
```

`knn_predict` is about as simple a model as you can build: predict a point by averaging its fifteen nearest neighbors in time. There is no mechanism in it, no claim about why nearby points should look alike, just persistence exploited directly. Cross-validate it three ways: the textbook random shuffle, a version that respects contiguous blocks of time, and a blocked version that also removes a buffer around each held-out block.

```{r}
#| label: cv-leakage-compare
baseline_rmse <- sqrt(mean((y - mean(y))^2))

# 1. ordinary random 10-fold CV
k <- 10
folds <- sample(rep(1:k, length.out = n_sim))
err_random <- numeric(n_sim)
for (f in 1:k) {
  test_idx  <- which(folds == f)
  train_idx <- setdiff(1:n_sim, test_idx)
  err_random[test_idx] <- (y[test_idx] -
    knn_predict(x[train_idx], y[train_idx], x[test_idx]))^2
}

# 2. blocked CV: contiguous folds, no shuffling
blockid <- cut(1:n_sim, k, labels = FALSE)
err_block <- numeric(n_sim)
for (f in 1:k) {
  test_idx  <- which(blockid == f)
  train_idx <- setdiff(1:n_sim, test_idx)
  err_block[test_idx] <- (y[test_idx] -
    knn_predict(x[train_idx], y[train_idx], x[test_idx]))^2
}

# 3. h-block CV: blocked, plus a buffer removed from training around each block
h <- 15
err_hblock <- numeric(n_sim)
for (f in 1:k) {
  test_idx <- which(blockid == f)
  buffer_idx <- unique(unlist(lapply(test_idx, function(i) max(1, i - h):min(n_sim, i + h))))
  train_idx <- setdiff(1:n_sim, buffer_idx)
  err_hblock[test_idx] <- (y[test_idx] -
    knn_predict(x[train_idx], y[train_idx], x[test_idx]))^2
}

c(baseline = baseline_rmse,
  random_kfold = sqrt(mean(err_random)),
  blocked = sqrt(mean(err_block)),
  h_block = sqrt(mean(err_hblock)))
```

Look at that comparison. The baseline, predicting nothing but the series mean, comes in at `r round(baseline_rmse, 2)`. Naive random k-fold CV reports an RMSE of `r round(sqrt(mean(err_random)), 2)`, a big improvement over baseline, which looks like skill. There is none. `x` is nothing but a row index; `y` has no relationship to it beyond its own persistence, which is exactly what the random fold structure leaks. Shuffle the rows, and most held-out points still have their immediate temporal neighbors sitting in the training set a few indices away, so a nearest-neighbor predictor effectively gets to peek at the answer. Respect the blocks of time instead, and the leak closes: blocked CV comes back at `r round(sqrt(mean(err_block)), 2)`, no better than guessing the mean. Add a buffer around each held-out block, the h-block fix [@Bergmeir2012], and it lands at `r round(sqrt(mean(err_hblock)), 2)`, confirming there was never anything here to find. The random fold did not measure the model. It measured how much information leaked across the shuffle.

## Rolling-Origin Validation

Once you accept that a time series cannot be shuffled, the alternative is to respect the one piece of structure that actually matters: order. *Forecasting: Principles and Practice* calls this time series cross-validation [@Hyndman2021fpp3]; you will also see it called rolling-origin or walk-forward validation elsewhere. Pick a minimum training window, fit, forecast one step ahead, score it, then slide the origin forward by one observation and repeat. Every point past the minimum window gets used as a test case exactly once, always predicted from data that came strictly before it. This is the single calibrate/verify split from the reconstruction chapter, repeated at every possible point along the record instead of one.

```{r}
#| label: cv-rolling-origin
#| message: false
coRiverFlow <- read_csv("data/colorado_flow.csv")
instrumental <- coRiverFlow |> filter(!is.na(ObsMAF)) |> arrange(Year)
n <- nrow(instrumental)

min_train <- 30
preds <- rep(NA_real_, n)
for (t in min_train:(n - 1)) {
  m <- lm(ObsMAF ~ Proxy, data = instrumental[1:t, ])
  preds[t + 1] <- predict(m, instrumental[t + 1, ])
}

rolling_rmse <- sqrt(mean((instrumental$ObsMAF - preds)^2, na.rm = TRUE))
rolling_rmse
```

That single number summarizes `r n - min_train` separate one-step forecasts, each made from a model that had never seen the year it was predicting. Compare it to the static split the reconstruction chapter actually used, calibrating through 1955 and verifying the rest in one block.

```{r}
#| label: cv-rolling-vs-static
calib <- instrumental |> filter(Year <= 1955)
verif <- instrumental |> filter(Year >= 1956)
static_rmse <- sqrt(mean((verif$ObsMAF -
  predict(lm(ObsMAF ~ Proxy, data = calib), verif))^2))
c(rolling_origin = rolling_rmse, static_split = static_rmse)
```

The two roughly agree, which is itself informative: the relationship's skill does not depend on which years happened to fall in the held-out block, it holds up wherever you put the wall. That is exactly the question the "Move the wall" exercise back in the reconstruction chapter asked you to check by hand at a couple of points. Rolling-origin validation is that exercise automated and exhaustive, run at every wall instead of two or three.

## Blocked Cross-Validation, and When You'd Use It

Rolling-origin validation only ever predicts forward, which is the right discipline when forecasting is the actual goal. But the reconstruction chapter was not forecasting the future. It was estimating a fixed relationship, between a proxy and a flow, assumed to hold the same way across the whole period, then applied somewhere else entirely, into years with no flow data at all. For that kind of problem strict forward-only validation is not required. What matters is that a held-out point's own temporal neighbors never sit in the training data. That is what blocked cross-validation, with a buffer if the residuals carry any memory, is for [@Bergmeir2012]. Drop a contiguous chunk, train on the rest, predict the chunk, rotate to the next chunk. It uses the data more fully than a single calibrate/verify split, without pretending the problem is a forecast it is not.

## A Surprising Exception

Here is the wrinkle that keeps this from being a blanket rule. Bergmeir, Hyndman, and Koo went back and asked exactly when ordinary, fold-shuffling cross-validation is valid for time series, and the answer is not "never" [@BergmeirHyndmanKoo2018]. For a model that is purely autoregressive and correctly specified, one where the errors really are independent once the model is fit, standard k-fold CV, shuffling included, holds up. The leakage in the demonstration above came from a model with nothing in it but unmodeled persistence, persistence the cross-validation scheme then let the model cheat with. Fit a model that actually accounts for that persistence, so the residuals are no longer autocorrelated, and there is no leftover memory for a shuffle to exploit.

That is the same diagnostic this book keeps returning to. What breaks cross-validation is not autocorrelation in the raw series, it is autocorrelation left over after the model has had its turn. Check the residuals, the way the regression and reconstruction chapters insist on, and you know which situation you are in before you pick a validation scheme.

## Where This Shows Up

The forecasting chapter's single backtest and the reconstruction chapter's single calibrate/verify split are both correct, minimal versions of validation, and a fine place to start. Reach for rolling-origin validation when you want to use a long record more fully and the model only predicts forward. Reach for blocked cross-validation, with a buffer, when you are estimating a relationship rather than forecasting and want more than one split's worth of evidence for it. And before reaching for either, look at the residuals. A model that has already absorbed the series' memory does not need any of this machinery to be cross-validated correctly. A model that has not will fail quietly, the way the nearest-neighbor demonstration did, looking its best exactly when you should trust it least.
