# The Frequency Domain

```{r}
#| label: freq-setup
#| echo: false
set.seed(872)
```

## Big Idea

Every chapter up to now has kept the same question in view: given what a series has done, what does that say about what it does next? That's a time-domain question, the right one for autocorrelation, ARMA, forecasting, and filtering. This chapter throws out the clock. Instead of asking when something happens, we ask how often it happens, and the payoff is a different picture of the same data: variance broken out by frequency instead of laid along a timeline.

It takes a minute to get used to. A squiggly line and a spike at some particular frequency don't look like the same object, but they carry the same information, just organized differently, the way a chord and a list of its individual notes describe the same sound. Once you can move between the two views, some questions that are nearly impossible to answer by eyeballing a time-domain plot become almost trivial. Is there an eleven-year cycle buried in three centuries of sunspot counts? Is there a signature of the Earth's own orbit hiding in five million years of sunlight? Both show up later in this chapter, and in both cases nobody could point to the cycle in the raw time-domain plot. The frequency domain finds it anyway.

## Packages

We'll build every periodogram in this chapter with base R's own `spectrum()` function, so nothing about a Fourier transform happens out of view. `tidyverse` does the wrangling and plotting, `PNWColors` supplies the figure colors, and `cowplot` stacks a time-domain plot on top of its frequency-domain twin. The wavelet material that used to close this chapter now lives in the aside right after it, which loads `waveslim` and `dplR` on its own.

```{r}
#| label: freq-packages
#| message: false
library(tidyverse)
library(PNWColors)
library(cowplot)

sfPal <- pnw_palette("Starfish", n = 5, type = "discrete")
```

## Same Data, Two Domains

Thus far we've looked at a time series as a squiggly line, some variable changing as a function of time, with the horizontal axis marking when things happen. That view is built for detecting trends, abrupt shifts, and autocorrelation, and it's the one this whole book has used up to here.

The frequency domain asks a different question: what cycles are embedded in this series? Instead of "when," it asks "how often," and it represents the series as a set of periodic components, the frequencies that make it up. The result is a plot of variance as a function of frequency: how much of the up-and-down in the data comes from a process operating at a fast timescale, and how much from a slow one.

This isn't just a mathematical trick. Plenty of environmental systems are driven by periodic forcing (daily, seasonal, tidal, orbital), and working in the frequency domain often gives a clearer view of what's driving them. It's also good at pulling a signal out of noise, since a cycle that's invisible against the day-to-day jitter of a time-domain plot can stand out sharply once its variance is concentrated in one narrow slice of frequency.

The tool behind all of this is spectral analysis: decomposing a signal into its component frequencies. The idea goes back to Joseph Fourier, who showed that any periodic signal can be written as a sum of sine and cosine waves, a Fourier series. Even signals that aren't obviously periodic can often be approximated this way. A time series that looks like noise may just be several overlapping cycles at different frequencies and amplitudes, stacked on top of each other.

The Fourier transform is the machinery that does the decomposing, and it's the same idea behind MP3 audio compression and JPEG images: represent something complicated as a sum of simple waves, then keep only the ones that matter. Here we'll use it to find the frequencies embedded in a time series and ask which are strong, which are weak, and which are just noise.

The payoff we're after is mechanism, not just description. When a peak in the frequency domain lines up with a known physical driver, whether El Niño, the solar cycle, or Milankovitch forcing, that's what turns a periodogram from a plot into evidence.

Here are two plots of the same signal, one in each domain.

```{r}
#| label: freq-two-domains
#| fig-height: 6
n <- 1000
tm <- 1:n
twoDomainSignal <- sin(2 * pi * tm / 50) + 0.5 * sin(2 * pi * tm / 200) + rnorm(n, sd = 0.5)
twoDomainDat <- tibble(time = tm, signal = twoDomainSignal)

pTime <- ggplot(twoDomainDat, aes(time, signal)) +
  geom_line(color = sfPal[1]) +
  labs(title = "Time Domain", x = "Time", y = "Signal") +
  theme_minimal()

twoDomainSpec <- spectrum(twoDomainSignal, plot = FALSE)
twoDomainSpecDat <- tibble(freq = twoDomainSpec$freq, spec = twoDomainSpec$spec)

pFreq <- ggplot(twoDomainSpecDat, aes(freq, spec)) +
  geom_line(color = sfPal[5]) +
  labs(title = "Frequency Domain", x = "Frequency (1 / time unit)", y = "Spectral Density") +
  theme_minimal()

plot_grid(pTime, pFreq, ncol = 1)
```

The two plots carry exactly the same information, just through a different lens.^[In theory the time and frequency domains are mathematically equivalent, and you can reconstruct one from the other with a full Fourier transform. In practice, smoothing, truncation, or sampling limits can obscure some details in one domain that the other shows plainly.] The time-domain plot shows the signal as it changes; the frequency-domain plot shows how much of the signal's variance sits at each frequency. You can see strong peaks around 0.02 (a 50-unit cycle) and 0.005 (a 200-unit cycle), exactly the two sine waves this signal was built from. That's spectral analysis working as advertised: it reveals periodicity that's actually there even when the time-domain plot just looks like noise.

## Can We Recover the Cycles We Built?

Simulated data are convenient for exactly this reason: you get to write the answer key before you look at the periodogram. We'll build a signal out of four sine waves with known periods, bury it in noise, and check whether `spectrum()` finds all four.

Here's the first one, a sine wave with a period of 250 (a frequency of 1/250 = 0.004), so over 1000 observations it completes four full cycles.

```{r}
#| label: freq-wave-250
waveDat <- tibble(tm = 1:1000, wav250 = 0.3 * sin(2 * pi / 250 * tm))

ggplot(waveDat, aes(tm, wav250)) +
  geom_line(color = sfPal[1]) +
  labs(x = "Time", y = "Signal",
       title = "Period 250, frequency = 0.004", subtitle = "Amplitude = 0.3") +
  theme_minimal()
```

Now three more, each with its own period and amplitude.

```{r}
#| label: freq-wave-others
waveDat <- waveDat |>
  mutate(wav50 = 0.75 * sin(2 * pi / 50 * tm),
         wav10 = 1.00 * sin(2 * pi / 10 * tm),
         wav5  = 0.50 * sin(2 * pi / 5  * tm))
```

```{r}
#| label: freq-wave-50-plot
ggplot(waveDat, aes(tm, wav50)) +
  geom_line(color = sfPal[2]) +
  labs(x = "Time", y = "Signal",
       title = "Period 50, frequency = 0.02", subtitle = "Amplitude = 0.75") +
  theme_minimal()
```

```{r}
#| label: freq-wave-10-plot
ggplot(waveDat, aes(tm, wav10)) +
  geom_line(color = sfPal[3]) +
  labs(x = "Time", y = "Signal",
       title = "Period 10, frequency = 0.1", subtitle = "Amplitude = 1") +
  theme_minimal()
```

```{r}
#| label: freq-wave-5-plot
ggplot(waveDat, aes(tm, wav5)) +
  geom_line(color = sfPal[4]) +
  labs(x = "Time", y = "Signal",
       title = "Period 5, frequency = 0.2", subtitle = "Amplitude = 0.5") +
  theme_minimal()
```

Sum all four and the individual waves blur into something that already looks a lot less tidy.

```{r}
#| label: freq-wave-sum
waveDat <- waveDat |> mutate(allWave = wav250 + wav50 + wav10 + wav5)

ggplot(waveDat, aes(tm, allWave)) +
  geom_line(color = sfPal[5]) +
  labs(x = "Time", y = "Signal", title = "Four frequencies combined") +
  theme_minimal()
```

Zoom in on a hundred-point stretch and the mix is even harder to read by eye.

```{r}
#| label: freq-wave-zoom
ggplot(waveDat, aes(tm, allWave)) +
  geom_line(color = sfPal[5]) +
  labs(x = "Time", y = "Signal", title = "Four frequencies combined") +
  lims(x = c(500, 600)) +
  theme_minimal()
```

Add some noise and this looks a lot like something you'd actually meet in the field: a little wiggle, a lot of repeating structure, none of it obvious from the plot alone.

```{r}
#| label: freq-wave-noisy
waveDat <- waveDat |> mutate(allWaveNoisy = allWave + rnorm(1000))

ggplot(waveDat, aes(tm, allWaveNoisy)) +
  geom_line(color = sfPal[5]) +
  labs(x = "Time", y = "Signal", title = "Four frequencies plus noise") +
  theme_minimal()
```

Now the question: can `spectrum()` recover the four cycles we know are buried in there? We'll look at the raw periodogram, unsmoothed, since we already know exactly what we're looking for.

```{r}
#| label: freq-wave-spectrum-raw
spectrum(waveDat$allWaveNoisy, log = "no")
```

By default the plot puts the spectral density on a log axis; it's turned off here, but run `spectrum(waveDat$allWaveNoisy)` yourself to see the log version, since that's what you'll get if you don't ask otherwise. The spectrum is scaled so that the area under the curve equals about half the variance of the series. `spectrum()` can also smooth the periodogram (the `span` argument), which matters once the data get noisy enough that every point looks like its own tiny spike; @Cowpertwait2009 works through why that smoothing is needed and what it costs. For this signal, though, the raw periodogram is exactly what we want, because we already know it should show four sharp peaks and nothing else.

```{r}
#| label: freq-wave-spectrum-reveal
waveSpec <- spectrum(waveDat$allWaveNoisy, plot = FALSE)
waveSpecDat <- tibble(freq = waveSpec$freq, spec = waveSpec$spec)

knownFreqs <- tibble(period = c(5, 10, 50, 250)) |> mutate(freq = 1 / period)

ggplot(waveSpecDat, aes(freq, spec)) +
  geom_line(color = "grey40") +
  geom_vline(data = knownFreqs, aes(xintercept = freq),
             linetype = "dashed", color = sfPal[5]) +
  labs(x = "Frequency", y = "Spectral density",
       title = "The periodogram recovers all four planted cycles",
       subtitle = "Dashed lines mark the frequencies we built in: periods 5, 10, 50, 250") +
  theme_minimal()
```

The four tallest peaks in that plot land right on the four dashed lines, periods 5, 10, 50, and 250, in that order by strength. Nothing was fit or tuned to make that happen. `spectrum()` just took the noisy sum and handed back the pieces it was built from, which is the whole promise of this chapter: a squiggly line can be a stack of cycles in disguise, and the frequency domain is how you find out which ones.

## An Observed Series: Sunspots

Simulated series are a good place to learn the mechanics, but let's point the same tool at data nobody built: the sunspot count, one of the longest continuously kept scientific records there is.

```{r}
#| label: freq-sunspot-annual-plot
data(sunspot.year)
plot(sunspot.year)
```

The cycling is obvious by eye, but how regular is it, really? Compute the periodogram and find out.

```{r}
#| label: freq-sunspot-annual-spec
sunspotAnnualSpec <- spectrum(sunspot.year, span = 5, plot = FALSE)
plot(sunspotAnnualSpec, log = "no", type = "h",
     xlab = "Frequency (cycles / yr)", ylab = "Spectral density")
```

There's a sharp spike around a frequency of 0.09. Since $f = 1/p$, that's a period of about `r round(1 / sunspotAnnualSpec$freq[which.max(sunspotAnnualSpec$spec)], 1)` years, right where the well-known eleven-year solar cycle should land, in a record that goes back to 1700.

Zoom in on the low end of the frequency axis and there's more going on.

```{r}
#| label: freq-sunspot-annual-zoom
plot(sunspotAnnualSpec, log = "no", type = "h",
     xlab = "Frequency (cycles / yr)", ylab = "Spectral density", xlim = c(0, 0.2))
```

Below the eleven-year spike there's a broad rise in power at much lower frequencies, periods out past fifty years, that never resolves into a single sharp peak the way the eleven-year cycle does. That's consistent with the Gleissberg cycle [@Peristykh2003], a slower, roughly eighty-to-one-hundred-year modulation of how strong each eleven-year cycle runs (some cycles stronger, some weaker, riding a slower wave of their own). `sunspot.year` only spans `r diff(range(time(sunspot.year)))` years, not even three full Gleissberg cycles, so what shows up here is a broad rise rather than a resolvable peak. That's worth remembering the next time a periodogram gives you a shapeless bump instead of a clean spike: sometimes that's a cycle the record is simply too short to pin down, not an absence of one.

You can also plot an object of class `spec` with `plot()`, as we just did (see `?plot.spec` for the full set of options). Here's a version built in `ggplot`, with a second x-axis for period, since after all this time I still think in periods, not frequencies.

```{r}
#| label: freq-sunspot-annual-ggplot
sunspotAnnualDat <- tibble(freq = sunspotAnnualSpec$freq, spec = sunspotAnnualSpec$spec)

freqBreaks <- pretty(sunspotAnnualDat$freq)
periodLabels <- round(1 / freqBreaks, 2)

ggplot(sunspotAnnualDat, aes(x = freq, xend = freq, y = 0, yend = spec)) +
  geom_segment() +
  scale_y_continuous(name = "Spectral density", expand = c(0, 0)) +
  scale_x_continuous(name = "Frequency (cycle / yr)",
                      sec.axis = sec_axis(~., breaks = freqBreaks, labels = periodLabels,
                                           name = "Period (yr / cycle)")) +
  theme_minimal()
```

Now the same data at monthly resolution instead of yearly.

```{r}
#| label: freq-sunspot-monthly-plot
data(sunspot.month)
plot(sunspot.month)
frequency(sunspot.month)
```

`sunspot.month` is the same underlying phenomenon, sampled twelve times a year instead of once; `frequency()` reports 12, the same start/frequency bookkeeping the intro chapter's `ts` bridge relies on. Think about what that change in sampling rate should do to the frequency axis before you look at the next plot.

```{r}
#| label: freq-sunspot-monthly-spec
sunspotMonthlySpec <- spectrum(sunspot.month, span = 5, plot = FALSE)
sunspotMonthlyDat <- tibble(freq = sunspotMonthlySpec$freq, spec = sunspotMonthlySpec$spec)

ggplot(sunspotMonthlyDat, aes(freq, spec)) +
  geom_line(color = sfPal[1]) +
  labs(x = "Frequency (cycle / yr)", y = "Spectral density") +
  theme_minimal()
```

The x-axis now reaches out to `r round(max(sunspotMonthlyDat$freq), 1)` cycles per year instead of stopping at 0.5. That's the Nyquist limit at work: the highest frequency a periodogram can resolve is always half the sampling rate, and twelve samples a year instead of one pushes the ceiling from 0.5 to 6. The eleven-year cycle is still the tallest thing in the plot (`r round(1 / sunspotMonthlyDat$freq[which.max(sunspotMonthlyDat$spec)], 1)` years, same story as the annual version), just squeezed into a sliver near zero on an axis that now stretches six times further. The extra room on that axis is empty for sunspot counts, since there's no meaningful monthly-scale cycle in how many spots the sun has, but the same Nyquist logic is exactly what would let a faster-sampled instrument (daily solar irradiance, say) pick up faster solar phenomena that yearly or monthly counts were never going to resolve either way.

## An Observed Series: Orbital Insolation and the Milankovitch Cycles

Now for the biggest jump in scale this book will make: from eleven years to five million. Many of you have run into the idea of orbital forcing through [Milankovitch cycles](https://en.wikipedia.org/wiki/Milankovitch_cycles), the slow changes in Earth's orbital eccentricity, its axial tilt (obliquity), and the wobble of that axis (precession). Each of those three orbital properties cycles on its own timescale, tens to hundreds of thousands of years, and together they're the leading explanation for the pacing of the ice ages.

`data/jul65N.csv` holds a reconstruction of July insolation (incoming solar energy, in W/m²) at 65°N latitude for the past five million years, from @BergerLoutre1991. Unlike every other dataset in this chapter, nothing here was measured. The whole series is calculated straight from celestial mechanics: given the Earth, the sun, and Newton's laws, you can work out exactly how much sunlight fell on 65°N in July at any point over that span. `kya` counts thousands of years relative to the present and runs negative into the past (0 is now, -5000 is five million years ago), which is why the plot below reads left-to-right as past-to-present without needing to be reversed. `W.per.m2` is the insolation itself.

```{r}
#| label: freq-milank-plot
#| message: false
insolation <- read_csv("data/jul65N.csv")

ggplot(insolation, aes(kya, W.per.m2)) +
  geom_line(color = sfPal[5]) +
  labs(x = "Thousands of years ago", y = expression(W/m^2),
       title = "July insolation at 65°N, past 5 million years") +
  theme_minimal()
```

It's a mess to eyeball. There's clearly some rhythm buried in there, but no single obvious period the way the eleven-year sunspot cycle jumps out of a time-domain plot. That's exactly the situation the frequency domain is built for.

```{r}
#| label: freq-milank-spec
insolationSpec <- spectrum(insolation$W.per.m2, span = 5, plot = FALSE)
peakFreq   <- insolationSpec$freq[which.max(insolationSpec$spec)]
peakPeriod <- 1 / peakFreq
peakPeriod
```

The tallest peak sits at a period of about `r round(peakPeriod, 1)` thousand years, squarely inside the precession band (roughly 19 to 24 thousand years, the two dominant periods produced by the wobble of Earth's axis interacting with its elliptical orbit). Plot the whole spectrum with the three Milankovitch bands marked and the rest of the picture fills in.

```{r}
#| label: freq-milank-bands-plot
milankovitchBands <- tibble(cycle = c("precession", "precession", "obliquity", "eccentricity"),
                             period = c(23, 19, 41, 100)) |>
  mutate(freq = 1 / period)

insolationSpecDat <- tibble(freq = insolationSpec$freq, spec = insolationSpec$spec)

ggplot(insolationSpecDat, aes(freq, spec)) +
  geom_line(color = "grey40") +
  geom_vline(data = milankovitchBands, aes(xintercept = freq, color = cycle),
             linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = c(precession = sfPal[1], obliquity = sfPal[3],
                                 eccentricity = sfPal[5])) +
  scale_x_continuous(limits = c(0, 0.06)) +
  labs(x = "Frequency (cycles / kyr)", y = "Spectral density", color = NULL,
       title = "The Milankovitch bands in July insolation at 65°N") +
  theme_minimal()
```

All three bands are there. Two close, tall peaks sit in the precession band, and a smaller peak lands right at the 41,000-year obliquity period. Eccentricity, at 100,000 years, is barely there at all: its peak is about `r round(insolationSpecDat$spec[which.min(abs(insolationSpecDat$freq - 1/100))] / max(insolationSpecDat$spec) * 100, 2)`% the height of the tallest peak in the plot, invisible on this linear scale. Every one of those numbers came out of celestial mechanics, not from fitting anything to the data. The frequency domain just found them.

Here's the twist, and it's the reason paleoclimatologists still argue about this. Eccentricity is by far the weakest signal in that plot, yet the last 800,000 years of ice-core and ocean-sediment records are dominated by a roughly 100,000-year cycle between glacial and interglacial periods, riding on the weakest of the three orbital bands rather than the strongest. If ice volume responded to insolation the way a simple linear filter would, the biggest swings ought to track precession, the tallest band here, not a peak too small to see. It doesn't work that way, and untangling why is known in the field as the 100-kyr problem: something in the ice-sheet or carbon-cycle system amplifies the weak eccentricity signal far out of proportion to its size, and there's no full consensus on what that something is. The periodogram hands you the forcing, clean and unambiguous. What the planet does with that forcing is still an open question, which is exactly the kind of gap between pattern and process this book keeps circling back to.

> A handful of sine waves, derived entirely from orbital geometry, reconstruct the pacing of the ice ages before a single ice core is ever drilled. That's the frequency domain working at the largest scale this book will touch.

## Wrapping Up

The frequency domain answers a question the time domain can't: not when does something happen, but how often, and how much of the total variance belongs to each timescale. We built that idea from scratch with four sine waves whose periods we already knew, buried them in noise, and watched `spectrum()` hand every one of them back. Then we pointed the same tool at data nobody built: three centuries of sunspot activity, and five million years of the Earth's own orbit. Both cycles were sitting there the whole time, waiting for the right axis to become visible.

One thing worth carrying forward: a periodogram assumes the frequency content of a series stays put for the whole record, whatever cycles are there at the beginning are assumed to still be there at the end. That's a stationarity assumption wearing a different costume, and it isn't automatically true of environmental data (a species' calling frequency shifting as its habitat warms, a tree's dominant growth cycle changing as it ages past its fastest-growing years). When that assumption breaks down, a plain periodogram will average two very different eras together and hand you a blurred picture of both. There's a fix, wavelets, and it's worth a look even though nothing in it will be graded: see [the wavelets aside](12xAsideWavelets.html) right after this chapter.

## Exercises

### Three Signals, Three Spectra

Generate a reasonably long white-noise series, an AR(1) series, and a random walk (`cumsum()` of white noise works fine). Compute and plot the periodogram of each. Do the shapes match what you'd have guessed beforehand? Hint: think about which of the three series counts as stationary in the sense the stationarity chapter defined, and what that might do to a tool that assumes a series's frequency content doesn't drift over the record.

### Build Your Own Cycles

Go back to the four-sine-wave demo and make it your own. Change an amplitude, add a fifth wave, or lay a linear trend on top of everything and see what happens to the periodogram. Guess first: where should a trend show up in the frequency domain? (Hint: a trend is about as low-frequency as a signal gets.) Then check `?spectrum` for its `detrend` argument before you decide your guess was right or wrong. Also try smoothing the periodogram with the `span` argument, and see how much detail you can smooth away before the four peaks blur into the noise floor.

## Further Reading

- @Cowpertwait2009, ch. 9, works through spectral analysis of AR(1) processes for both positive and negative $\phi$, a good companion if you want the fuller derivation of what a periodogram is actually estimating, and what its smoothing (`span`) is doing for you.
- @BergerLoutre1991 is the original orbital solution behind `jul65N.csv`, computed well before satellite-era climate science and still a standard reference in paleoclimatology.
- @Peristykh2003 documents the Gleissberg cycle using cosmogenic-isotope records reaching back roughly 12,000 years, far past what the annual sunspot count alone can show.
- [The wavelets aside](12xAsideWavelets.html) picks up where this chapter's Wrapping Up leaves off: what to do when a series's frequency content changes over time.
