# Wavelets

```{r}
#| label: wave-setup
#| echo: false
set.seed(872)
```

## Big Idea

The last chapter ended on a warning: a periodogram assumes a series's frequency content holds steady for the whole record, whatever cycles are there at the beginning are assumed to still be there at the end. Plenty of environmental signals don't cooperate with that assumption. An animal's activity cycle can shift with the seasons. A lake's plankton bloom can drift earlier as the water warms. A tree's dominant growth rhythm can change once it ages past its fastest-growing years. A climate teleconnection's period can wander across decades. A periodogram will still run on data like that, and it will still hand back an answer, but the answer is a single spectrum averaged over the whole record, blind to when anything happened. Wavelets are the tool built to keep both questions open at once: what frequency, and when.

## Packages

`dplR` supplies `morlet()` and `wavelet.plot()` for the continuous wavelet transform, built on the algorithm from @TorrenceCompo1998. `waveslim` supplies `mra()` for the discrete version. `tidyverse` and `PNWColors` handle the planted demo.

```{r}
#| label: wave-packages
#| message: false
library(tidyverse)
library(PNWColors)
library(dplR)
library(waveslim)

sfPal <- pnw_palette("Starfish", n = 5, type = "discrete")
```

## Can a Periodogram Tell You When?

Build a signal that changes its mind partway through: a period-10 cycle for the first 200 points, a period-30 cycle for the next 200, plus a little noise so it isn't too tidy.

```{r}
#| label: wave-switch-build
n <- 400
tm <- 1:n
switchPoint <- 200

switchDat <- tibble(tm = tm,
                     signal = ifelse(tm <= switchPoint,
                                      sin(2 * pi * tm / 10),
                                      sin(2 * pi * tm / 30)) + 
                      rnorm(n, sd = 0.75))
                      #arima.sim(model=list(ar=0.1),n=n))

ggplot(switchDat, aes(tm, signal)) +
  geom_line(color = sfPal[1]) +
  geom_vline(xintercept = switchPoint, linetype = "dashed", color = "grey40") +
  labs(x = "Time", y = "Signal", title = "A period-10 cycle, then a period-30 cycle") +
  theme_minimal()
```

The switch is obvious by eye. Now run it through the same tool the last chapter built the whole book's frequency intuition on.

```{r}
#| label: wave-switch-periodogram
switchSpec <- spectrum(switchDat$signal, plot = FALSE)
switchSpecDat <- tibble(freq = switchSpec$freq, spec = switchSpec$spec)

ggplot(switchSpecDat, aes(freq, spec)) +
  geom_line(color = sfPal[5]) +
  geom_vline(xintercept = c(1 / 10, 1 / 30), linetype = "dashed", color = "grey40") +
  labs(x = "Frequency", y = "Spectral density", title = "The periodogram finds both periods") +
  theme_minimal()
```

Both periods show up, cleanly, right where they should. As far as identifying *what's* in the signal, the periodogram did its job. But look at the plot again: there's no time axis on it anywhere. It cannot tell you that period 10 ran first and period 30 ran second, or that they never overlapped, or where the handoff happened, because none of that information is in a periodogram's design. It reports how much of the whole record's variance belongs to each frequency, full stop. Feed it a series whose rhythm changes and it will still answer confidently, just not the question you actually wanted asked.

## Seeing the Switch: A Continuous Wavelet Transform

A continuous wavelet transform (CWT) keeps time on one axis and period on the other, with color showing how much power a given period carries at a given moment. Run it on the exact same series.

```{r}
#| label: wave-switch-cwt
#| fig-height: 6
switchCwt <- morlet(y1 = switchDat$signal, x1 = switchDat$tm, p2 = 8, dj = 0.1, siglvl = 0.99)
wavelet.plot(switchCwt, reverse.y = TRUE, crn.lab = "Signal", crn.col = "grey50")
```

Now the switch is right there in the picture: a band of power around period 10 that holds from the start and fades out near $t=200$, and a second band around period 30 that starts almost exactly where the first one stops. Nothing about *what* frequencies are present changed from the periodogram's answer. What changed is that this plot also answers *when*.

The hatched regions at the edges and at long periods are the cone of influence: past that line, the wavelet's window runs off the end of the record, and the estimate there is built on less evidence than it looks like it is. That's the same edge problem the filters chapter found with loess, drawn explicitly onto the plot instead of left for you to notice on your own.

## An Observed Series: A Tree-Ring Chronology

Measured records rarely hand you a clean, one-time switch. What they give you instead is a cycle that comes and goes, gets stronger, gets weaker, and maybe reappears somewhere else in the record. Here's the same `co021` tree-ring data the filters chapter used, this time averaged across many trees into a single site chronology rather than looked at core by core.

```{r}
#| label: wave-chron
data(co021)
co021_rwi <- detrend(co021, method = "AgeDepSpline")
co021_crn <- chron(co021_rwi)

chronVal <- co021_crn[, 1]
chronYrs <- as.numeric(time(co021_crn))

plot(co021_crn, add.spline = TRUE, nyrs = 64)
```

```{r}
#| label: wave-tree-cwt
#| fig-height: 6
treeCwt <- morlet(y1 = chronVal, x1 = chronYrs, p2 = 8, dj = 0.1, siglvl = 0.99)
wavelet.plot(treeCwt, crn.col = "grey50",reverse.y = TRUE)
```

Read this one like a topographic map of frequency content over time. A horizontal band means that period stayed active for as long as the band runs; a band that appears, breaks up, and reappears somewhere else is exactly the kind of feature a single periodogram of the whole 788-year record would average into invisibility. Whatever produces the multi-decadal wiggle in a tree's growth (drought cycles, temperature regimes, competition among neighboring trees) apparently doesn't run at a constant strength the whole time, and this is the tool that lets you say so.

## A Discrete Decomposition

The continuous transform makes a picture. The discrete version, a multiresolution analysis (MRA), does something more like the by-hand decomposition from the earliest chapters: it splits the series into additive pieces, one per timescale, so that summing every piece back together reconstructs the original series exactly.

```{r}
#| label: wave-dwt
nScales <- trunc(log(length(chronYrs)) / log(2)) - 1
dwtOut <- mra(chronVal, wf = "la8", J = nScales, method = "modwt", boundary = "periodic")

scaleLabels <- paste0(2^(1:nScales), " yrs")
dwtScaled <- scale(as.data.frame(dwtOut))
```

```{r}
#| label: wave-dwt-plot
#| fig-height: 7
par(mar = c(3, 2, 2, 2), mgp = c(1.25, 0.25, 0), tcl = 0.5, xaxs = "i", yaxs = "i")
plot(chronYrs, rep(1, length(chronYrs)), type = "n", axes = FALSE, ylab = "", xlab = "",
     ylim = c(-3, 38))
title(main = "Multiresolution decomposition of the chronology", line = 0.75)
axis(side = 1)
mtext("Years", side = 1, line = 1.25)

offset <- 0
for (i in nScales:1) {
  lines(chronYrs, dwtScaled[, i] + offset)
  abline(h = offset, lty = "dashed")
  mtext(colnames(dwtScaled)[i], side = 2, at = offset, line = 0)
  mtext(scaleLabels[i], side = 4, at = offset, line = 0)
  offset <- offset + 5
}
box()
```

Each row is one timescale, named on the left (`D1` through `D8`, plus a smooth `S8` at the bottom) and labeled on the right in years. The same idea applies to the switching demo above: a detail level tuned to period-10-ish variability should carry most of its variance in the first half of the record, and a level tuned to period-30-ish variability should carry most of its variance in the second half.

```{r}
#| label: wave-switch-dwt-check
switchScales <- trunc(log(n) / log(2)) - 1
switchDwt <- mra(switchDat$signal, wf = "la8", J = switchScales, method = "modwt", boundary = "periodic")

tibble(level = names(switchDwt),
       var_first_half  = map_dbl(switchDwt, ~ var(.x[1:switchPoint])),
       var_second_half = map_dbl(switchDwt, ~ var(.x[(switchPoint + 1):n]))) |>
  mutate(across(starts_with("var"), \(x) round(x, 3)))
```

`D3`, the detail level tuned closest to a period-10 cycle, carries far more variance in the first half than the second. `D4`, tuned closer to period 30, does the opposite. Same conclusion as the heat map, reached by a completely different route: split the series into timescales, and the timescales themselves tell you where the switch happened.

## Wrapping Up

A wavelet transform buys you something a periodogram can't give you (localization in time) but nothing is free: sharpening the time axis costs you some sharpness on the frequency axis, so a wavelet's estimate of "what period" is fuzzier than a periodogram's. That trade is worth making the moment you suspect a series's rhythm isn't constant, which for environmental data is more often than not.

This closes out the book's tour of the frequency domain. The early chapters lived entirely in the time domain, tracking a series left to right and asking what today has to do with yesterday. The frequency chapter flipped that around and asked what cycles a series is built from, no clock required. This chapter split the difference: time and frequency, together, on one plot. Between the three views, you have three lenses to reach for on whatever series lands on your desk next, and knowing which one fits the question is most of the job.

## Exercises

### Move the Switchpoint

Rebuild the planted demo with the switch happening somewhere other than the midpoint (try `switchPoint <- 100` or `switchPoint <- 320`), or add a second switch back to the original period so the series goes period-10, period-30, period-10 again. Does the wavelet plot still localize the change cleanly? What happens to the cone of influence near a switch that sits close to either end of the record?

### Confirm It With Numbers

The `wave-switch-dwt-check` chunk above compared the variance of each MRA detail level across the two halves of the planted demo for one pair of periods. Redo that comparison for the tree-ring chronology instead: pick a detail level from the `wave-dwt` output, split the 788-year record into an early half and a late half, and compare variances the same way. Does the split match anything visible in the continuous wavelet plot above it?

## Further Reading

- @TorrenceCompo1998 is the paper `dplR`'s `morlet()` and `wavelet.plot()` implement, written for exactly this audience: readers who want to use a wavelet transform without first sitting through the full functional-analysis derivation.
- @PercivalWalden2000 is the fuller, more rigorous treatment behind `waveslim`'s `mra()` and the maximal overlap discrete wavelet transform.
